Sparse spectral and $p$-finite element methods for partial differential equations on disk-slices and trapeziums

### 2019-04-11 papers/halfdisk/halfdisk.tex
Changed naming convention for the OPs

\newcommand{\Pnk}{P_{n,k}}
\newcommand{\Pab}{P^{(a,b)}}
\newcommand{\Pnkab}{P_{n,k}^{(a,b)}}
\newcommand{\Wab}{{W^{(a,b)}}}
\newcommand{\Pmjab}{P_{m,j}^{(a,b)}}
\newcommand{\alphaab}{\alpha^{(a,b)}}
\newcommand{\betaab}{\beta^{(a,b)}}
\newcommand{\bigPab}{\bigP^{(a,b)}}
\newcommand{\Dnt}{D^\top_n}
\newcommand{\Wii}{W^{(1,1)}}
\newcommand{\Pii}{P^{(1,1)}}
\newcommand{\bigPii}{{\mathbb{P}^{(1,1)}}}
\newcommand{\Poo}{P^{(0,0)}}
\newcommand{\bigPoo}{{\mathbb{P}^{(0,0)}}}
\newcommand{\Poi}{P^{(0,1)}}
\newcommand{\Pio}{P^{(1,0)}}
\newcommand{\dx}{\frac{\partial}{\partial x}}
\newcommand{\dy}{\frac{\partial}{\partial y}}
\newcommand{\laplacewii}{\Delta_W^{(1,1)\to(1,1)}}
\newcommand{\laplacewtt}{\Delta_W^{(2,2)\to(0,0)}}
\newcommand{\laplaceoo}{\Delta^{(0,0)\to(2,2)}}
\newcommand{\biharmonic}{_2\Delta_W^{(2,2)\to(2,2)}}

@@

We show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that  generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations \cite{olver2013fast} and its analogue on the triangle \cite{olver2018recurrence,olver2019triangle}.  On the half-disk the family of weights we consider are of the form
$$
W^{(a,b)}(x,y) = x^a (1-x-y)^b
$$
with corresponding OPs denoted $P_{n,k}^{(a,b)}(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is,
$$
P_{n,k}^{(a,b)}(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
$$
where $C_{n,k} \neq 0$ and ``lower order terms'' can  include degree $n$ polynomials with fewer $y$ terms.

@@

In this section we outline the construction and some basic properties of $P_{n,k}^{(a,b)}(x,y)$. The symmetry in the weight allows us to express the polynomials in terms of 1D OPs, and deduce certain properties such as recurrence relationships.

@@

\begin{proposition}
Let \(w_1 : (\alpha,\beta) \: \to \R\), \(w_2 : (\gamma,\delta) \: \to \R\) be weight functions with \(\alpha,\beta,\gamma,\delta \in \R\), and let \(\rho \: : \: (\alpha,\beta) \: \to (0,\infty)\) be such that either:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(w_2\) is an even function.
\end{enumerate}
$\forall$, $n = 0,1,2,\dots, $ let $\{p_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k+1} w_1(x)$ where $0 \le k \le n$, and $\{q_{n}\}$ be polynomials orthogonal with respect to the weight $w_2(x)$. Then the 2D polynomials defined on $\Omega$
$$
\Pnk(x,y) := p_{n-k,k}(x) \: \rho(x)^k \: q_k\fpr(\frac{y}{\rho(x)}) \qquad\hbox{for} \qquad 0 \le k \le n, \: n = 0,1,2,\dots
$$
are orthogonal polynomials with respect to the weight \(W(x,y) := w_1(x) \: w_2(\frac{y}{\rho(x)}) \) on $\Omega$. 
\end{proposition}
\begin{proof}
See \cite[p55--56]{dunkl2014orthogonal}.
\end{proof}

We can generalise this for our purposes in the following proposition.
\begin{proposition}
Let
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
with $\rho : (\alpha, \beta) \to \R$. Further, let $w^{(a,b,c)}_H(x)$ and $w^{(a,b)}_P(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align}
\begin{cases}
w^{(a,b,c)}_H(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
w^{(a,b)}_P(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align}
where the three-parameter family of orthogonal polynomials on $[\alpha,\beta]$ given by $\{H_n^{(a,b,c)}\}$ are orthogonal with respect to $w^{(a,b,c)}_H(x)$, and the two-parameter family of orthogonal polynomials on $[\gamma,\delta]$ given by $\{P_n^{(a,b)}\}$ are orthogonal with respect to $w^{(a,b)}_P(x)$.

Assume that either of the following hold:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(w^{(a,b)}_P\) is an even function (i.e. $a = b$, and we can hence denote the weight as $w^{(a)}_P(x) = (\delta-x^2)^a$).
\end{enumerate}
Then the four-parameter 2D polynomials defined on $\Omega$ given by
\begin{align}
	\Pnk^{(a,b,c,d)}(x,y) := H_{n-k}^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: P_k^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
are orthogonal polynomials with respect to the weight 
\begin{align}
	W^{(a,b,c,d)}(x,y) := w^{(a,b,c+d)}_H(x) \: w_P^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega. 
\end{align}
\end{proposition}
\begin{proof}
See \cite[p55--56]{dunkl2014orthogonal}.
\end{proof}

For the half disk, the weight $W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b$ results from setting:
\begin{align}
\begin{cases}
(\alpha,\beta) &:= (0.1) \\
(\gamma,\delta) &:= (-1,1) \\
\rho(x) &:= (1-x^2)^{\half} \\
w_1(x) &:= w_H^{(a,b)}(x) := x^a \: (1-x^2)^b \\
w_2(x) &:= w_P^{(b)}(x) := (1-x^2)^b = (1-x)^b \: (1+x)^b,
\end{cases}
\end{align}
Note here the we have made the adjustment that $w^{(a,b,c)}_H(x) = (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{2c}$ and simply set the first parameter to zero and removed it for the specific half-disk case. This is for ease of writing and computation, by ensuring we reduce the need to incorporate unnecessary parameters and instead yield a two-parameter family of 2D orthogonal polynomials.

The weight $w_2(x)$ is an ultraspherical weight, and the corresponding OPs are   the Jacobi polynomials  \(\{P_n^{(b, b)}\}\). The weight $w_1(x)$ is non-classical (it is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}) and we introduce the notation \(\{H_n^{(a, b)}\}\) for the orthonormal polynomials. Thus we arrive at  a two parameter family of 2D orthogonal polynomials\sotodo{We should probably use a different letter than $P$ to avoid confusion with Jacobi polynomials}
\ \(\{\Pnkab\}\) given by, for \(0 \le k \le n, \: n = 0,1,2,\dots,\)
\begin{align}\label{eq:diskpolys}
 \Pnk^{(a,b)}(x,y) := H_{n-k}^{(a, b+k+\half)}(x) \: \rho(x)^k \: P_k^{(b,b)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
orthogonal on \(\Omega\) with respect to the weight
\begin{align}
W^{(a,b)}(x,y) = w_H^{(a,b)}(x) \: w_P^{(b)}(\frac{y}{\rho(x)}) = x^a \: (1-x^2-y^2)^b.
\end{align}

@@

We can express the three-term recurrences associated to $H_n^{(a,b)}$ and $P_n^{(b,b)}$ as
\begin{align}
x H_n^{(a,b)}(x) &= \beta_n^{(a,b)} H_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} H_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} H_{n-1}^{(a,b)}(x) 
\label{eqn:Hrecurrence} \\
y P_n^{(b,b)}(y) &= \delta_n^{(b)} P_{n+1}^{(b,b)}(y) + \gamma_n^{(b)} P_n^{(b,b)}(y) + \delta_{n-1}^{(b)} P_{n-1}^{(b,b)}(y),
\end{align}
where we note that \(\gamma_n^{(b)} = 0\), for \( n = 0,1,2,\dots\). We can use these to determine the 2D recurences for $P_{n,k}^{(a,b)}(x,y)$:

\begin{lemma}
$\Pnk^{(a,b)}(x,y)$ satisfy the following 3-term recurrences:
\begin{align}
x \Pnk^{(a,b)}(x,y) &= \alphaab_{n,k,1} \: P_{n-1, k}^{(a,b)}(x, y) + \alphaab_{n,k,2} \: P_{n, k}^{(a,b)}(x, y) + \alphaab_{n+1,k,1} \: P_{n+1, k}^{(a,b)}(x, y), \\
y \Pnk^{(a,b)}(x,y) &= \betaab_{n,k,1} \: P_{n-1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,2} \: P_{n-1, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,3} \: P_{n, k-1}^{(a,b)}(x, y) + \betaab_{n,k,4} \: P_{n, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,5} \: P_{n+1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,6} \: P_{n+1, k+1}^{(a,b)}(x, y),
\end{align}
for \((x,y) \in \Omega\), where
\begin{align}
\alphaab_{n,k,1} &:= \beta_{n-k-1}^{(a, b+k+\half)} \\
\alphaab_{n,k,2} &:= \alpha_{n-k}^{(a, b+k+\half)} \\
\betaab_{n,k,1} &:= {\delta_{k-1}^{(b)}}{|| P_{n-1,k-1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<H_{n-k}^{(a, b+k+1/2)}, H_{n-k}^{(a, b+k-1/2)}>_{w_H^{(a, b+k+1/2)}} \\
\betaab_{n,k,2} &:= {\delta_{k}^{(b)}}{|| P_{n-1,k+1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<H_{n-k}^{(a, b+k+1/2)}, H_{n-k-2}^{(a, b+k+{3/2})}>_{w_H^{(a, b+k+{3/2})}} \\
\betaab_{n,k,3} &:= {\delta_{k-1}^{(b)}}{|| P_{n,k-1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<H_{n-k}^{(a, b+k+1/2)}, H_{n-k+1}^{(a, b+k-1/2)}>_{w_H^{(a, b+k+1/2)}} \\
\betaab_{n,k,4} &:= {\delta_{k}^{(b)}}{|| P_{n,k+1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<H_{n-k}^{(a, b+k+1/2)}, H_{n-k-1}^{(a, b+k+{3/2})}>_{w_H^{(a, b+k+{3/2})}} \\
\betaab_{n,k,5} &:= {\delta_{k-1}^{(b)}}{|| P_{n+1,k-1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<H_{n-k}^{(a, b+k+1/2)}, H_{n-k+2}^{(a, b+k-1/2)}>_{w_H^{(a, b+k+1/2)}} \\
\betaab_{n,k,6} &:= {\delta_{k}^{(b)}}{|| P_{n+1,k+1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<H_{n-k}^{(a, b+k+1/2)}, H_{n-k}^{(a, b+k+{3/2})}>_{w_H^{(a, b+k+{3/2})}}. 
\end{align}
where $|| p(x,y) ||_{W^{(a,b)}}^2 := \int_\Omega p(x,y)^2 \: \Wab(x,y) \: d\mathbf{x}$ and $\ip<p(x), q(x)>_{w_H^{(a, b)}} := \int_\alpha^\beta p(x) \: q(x) \: x^a \: (1-x^2)^b \: dx$, i.e the 1D inner product using the $\{H^{(a,b)}_n\}$ family weight of $w_H^{(a, b)}$.

@@

\begin{proof}
Note that \(\ip<y \Pnkab, P_{m,j}^{(a,b)}>_\Wab = 0\) for \(m < n-1\) (and for \(m > n+2\)). Thus for \(m = n-1, n, n+1,\) \(j = 0,\dots,m:\)
\begin{align}
\ip<y \Pnkab, P_{m,j}^{(a,b)}>_\Wab &=  \iint_\Omega \Pnkab(x,y) \: \Pmjab(x,y) \: y \: \Wab(x,y) \: dy \: dx \\
&= \Big( \int^1_0 H_{n-k}^{(a, b+k+\half)}(s) \: H_{m-j}^{(a, b+j+\half)}(s) \: s^a \: \rho(s)^{2b+k+j+2} \: ds \Big) \nonumber \\
& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: (1-t^2)^b \: dt \Big) \\
&= \begin{cases}
    	\delta_k^{(b)} \: \ip<H_{n-k}^{(a, b+k+\half)}, H_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{w_H^{(a, b+k+\frac{3}{2})}} \quad& \text{if } j = k+1 \\
	\delta_{k-1}^{(b)} \: \ip<H_{n-k}^{(a, b+k+\half)}, H_{m-k+1}^{(a, b+k-\half)}>_{w_H^{(a, b+k+\half)}} \quad& \text{if } j = k-1 \\
	0 & \text{otherwise}
      \end{cases}.
\end{align}

@@

Three-term recurrences lead to Jacobi operators that correspond to multiplication by $x$ and $y$. Define, for $n=0,1,2,\dots$: 
\begin{align}
\bigPab_n := \begin{pmatrix}
		P^{(a,b)}_{n,-n}(x,y) \\
		\vdots \\
		P^{(a,b)}_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1}, 
\quad \quad 
\bigPab := \begin{pmatrix}
		\bigPab_0 \\
		\bigPab_1 \\
		\bigPab_2 \\
		\vdots \\
	\end{pmatrix}
\end{align}
and set $J_x^{(a,b)}, J_y^{(a,b)}$ as the Jacobi matrices corresponding to
\begin{align}
J_x^{(a,b)} \: \bigPab(x,y) = x \: \bigPab(x,y), \quad J_y^{(a,b)} \: \bigPab(x,y) = y \: \bigPab(x,y).
\label{eqn:jacobimatricesdefinition}
\end{align}
The matrices $J_x^{(a,b)}, J_y^{(a,b)}$ act on the coefficients vector of a functions expansion in the $\{\Pnkab\}$ basis. For example, let $a, b$ be general parameters and a function $f(x,y)$ defined on $\Omega$ be approximated by its expansion $f(x,y) = \bigPab(x,y)^\top \mathbf{f}$. Then $x \: f(x,y)$ is approximated by $\bigPab(x,y)^\top {J_x^{(a,b)\top}} \mathbf{f}$. In other words, ${J_x^{(a,b)\top}} \mathbf{f}$ is the coefficients vector for the expansion of the function $(x,y) \to x \: f(x,y)$ in the  $\{\Pnkab\}$ basis.

@@

We can then obtain the matrix operator for the Laplacian \(\Delta\), that will take us from coefficients for expansion in the weighted space
$$
\bigWii(x,y) = \Wii(x,y) \: \bigPii(x,y)
$$
to coefficients in the non-weighted space $\bigPii(x,y)$. Note that this construction will ensure the imposition of the Dirichlet zero boundary conditions on $\Omega$. The matrix operator for the Laplacian we denote $\laplacewii$ acting on the coefficients vector is then given by
\begin{align}
    \laplacewii := D_x^{(0,0)} \: W_x^{(1,1)} + T^{(0,1)\to(1,1)} \: D_y^{(0,0)} \: T_W^{(1,0)\to(0,0)} \: W_y^{(1,1)}.
\end{align}
Importantly, this operator will have Banded-Block-Banded structure, and hence will be sparse, as seen in Figure \ref{fig:sparsity}.

>>>>

\newcommand{\hdop}{H}
\newcommand{\bighdop}{\mathbb{\hdop}}
\newcommand{\hdopnk}{\hdop_{n,k}}
\newcommand{\hdopab}{\hdop^{(a,b)}}
\newcommand{\hdopnkab}{\hdop_{n,k}^{(a,b)}}
\newcommand{\Wab}{{W^{(a,b)}}}
\newcommand{\hdopmj}{\hdop_{m,j}}
\newcommand{\hdopmjab}{\hdop_{m,j}^{(a,b)}}
\newcommand{\alphaab}{\alpha^{(a,b)}}
\newcommand{\betaab}{\beta^{(a,b)}}
\newcommand{\bighdopab}{\bighdop^{(a,b)}}
\newcommand{\Dnt}{D^\top_n}
\newcommand{\Wii}{W^{(1,1)}}
\newcommand{\hdopii}{\hdop^{(1,1)}}
\newcommand{\bighdopii}{{\mathbb{\hdop}^{(1,1)}}}
\newcommand{\hdopoo}{\hdop^{(0,0)}}
\newcommand{\bighdopoo}{{\mathbb{\hdop}^{(0,0)}}}
\newcommand{\hdopoi}{\hdop^{(0,1)}}
\newcommand{\hdopio}{\hdop^{(1,0)}}
\newcommand{\genjac}{R}
\newcommand{\genjacnmk}{\genjac_{n-k}}
\newcommand{\genjacmmj}{\genjac_{m-j}}
\newcommand{\genjacw}{w_\genjac}

\newcommand{\dx}{\frac{\partial}{\partial x}}
\newcommand{\dy}{\frac{\partial}{\partial y}}
\newcommand{\laplacewii}{\Delta_W^{(1,1)\to(1,1)}}
\newcommand{\laplacewtt}{\Delta_W^{(2,2)\to(0,0)}}
\newcommand{\laplaceoo}{\Delta^{(0,0)\to(2,2)}}
\newcommand{\biharmonic}{_2\Delta_W^{(2,2)\to(2,2)}}

@@

We show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that  generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations \cite{olver2013fast} and its analogue on the triangle \cite{olver2018recurrence,olver2019triangle}.  On the half-disk the family of weights we consider are of the form
$$
W^{(a,b)}(x,y) = x^a (1-x-y)^b
$$
with corresponding OPs denoted $\hdopnkab(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is,
$$
\hdopnkab(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
$$
where $C_{n,k} \neq 0$ and ``lower order terms'' can  include degree $n$ polynomials with fewer $y$ terms.

@@

In this section we outline the construction and some basic properties of $\hdopnkab(x,y)$. The symmetry in the weight allows us to express the polynomials in terms of 1D OPs, and deduce certain properties such as recurrence relationships.

@@

\begin{proposition}
Let \(w_1 : (\alpha,\beta) \: \to \R\), \(w_2 : (\gamma,\delta) \: \to \R\) be weight functions with \(\alpha,\beta,\gamma,\delta \in \R\), and let \(\rho \: : \: (\alpha,\beta) \: \to (0,\infty)\) be such that either:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(w_2\) is an even function.
\end{enumerate}
$\forall$, $n = 0,1,2,\dots, $ let $\{p_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k+1} w_1(x)$ where $0 \le k \le n$, and $\{q_{n}\}$ be polynomials orthogonal with respect to the weight $w_2(x)$. Then the 2D polynomials defined on $\Omega$
$$
\hdopnk(x,y) := p_{n-k,k}(x) \: \rho(x)^k \: q_k\fpr(\frac{y}{\rho(x)}) \qquad\hbox{for} \qquad 0 \le k \le n, \: n = 0,1,2,\dots
$$
are orthogonal polynomials with respect to the weight \(W(x,y) := w_1(x) \: w_2(\frac{y}{\rho(x)}) \) on $\Omega$. 
\end{proposition}
\begin{proof}
See \cite[p55--56]{dunkl2014orthogonal}.
\end{proof}

We can generalise this for our purposes in the following proposition.
\begin{proposition}
Let
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
with $\rho : (\alpha, \beta) \to \R$. Further, let $\genjacw^{(a,b,c)}(x)$ and $w^{(a,b)}_P(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align}
\begin{cases}
\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
w^{(a,b)}_P(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align}
where the three-parameter family of orthogonal polynomials on $[\alpha,\beta]$ given by $\{\genjac_n^{(a,b,c)}\}$ are orthogonal with respect to $\genjacw^{(a,b,c)}(x)$, and the two-parameter family of orthogonal polynomials on $[\gamma,\delta]$ given by $\{P_n^{(a,b)}\}$ are orthogonal with respect to $w^{(a,b)}_P(x)$.

Assume that either of the following hold:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(w^{(a,b)}_P\) is an even function (i.e. $a = b$, and we can hence denote the weight as $w^{(a)}_P(x) = (\delta-x^2)^a$).
\end{enumerate}
Then the four-parameter 2D polynomials defined on $\Omega$ given by
\begin{align}
	\hdopnk^{(a,b,c,d)}(x,y) := \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: P_k^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
are orthogonal polynomials with respect to the weight 
\begin{align}
	W^{(a,b,c,d)}(x,y) := \genjacw^{(a,b,c+d)}(x) \: w_P^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega. 
\end{align}
\end{proposition}
\begin{proof}
See \cite[p55--56]{dunkl2014orthogonal}.
\end{proof}

For the half disk, the weight $W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b$ results from setting:
\begin{align}
\begin{cases}
(\alpha,\beta) &:= (0.1) \\
(\gamma,\delta) &:= (-1,1) \\
\rho(x) &:= (1-x^2)^{\half} \\
w_1(x) &:= \genjacw^{(a,b)}(x) := x^a \: (1-x^2)^b \\
w_2(x) &:= w_P^{(b)}(x) := (1-x^2)^b = (1-x)^b \: (1+x)^b,
\end{cases}
\end{align}
Note here the we have made the adjustment that $\genjacw^{(a,b,c)}(x) = (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{2c}$ and simply set the first parameter to zero and removed it for the specific half-disk case. This is for ease of writing and computation, by ensuring we reduce the need to incorporate unnecessary parameters and instead yield a two-parameter family of 2D orthogonal polynomials.

The weight $w_2(x)$ is an ultraspherical weight, and the corresponding OPs are   the Jacobi polynomials  \(\{P_n^{(b, b)}\}\). The weight $w_1(x)$ is non-classical (it is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}) and we introduce the notation \(\{\genjac_n^{(a, b)}\}\) for the orthonormal polynomials. Thus we arrive at  a two parameter family of 2D orthogonal polynomials\sotodo{We should probably use a different letter than $P$ to avoid confusion with Jacobi polynomials}
\ \(\{\hdopnkab\}\) given by, for \(0 \le k \le n, \: n = 0,1,2,\dots,\)
\begin{align}\label{eq:diskpolys}
 \hdopnk^{(a,b)}(x,y) := \genjacnmk^{(a, b+k+\half)}(x) \: \rho(x)^k \: P_k^{(b,b)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
orthogonal on \(\Omega\) with respect to the weight
\begin{align}
W^{(a,b)}(x,y) = \genjacw^{(a,b)}(x) \: w_P^{(b)}(\frac{y}{\rho(x)}) = x^a \: (1-x^2-y^2)^b.
\end{align}

@@

We can express the three-term recurrences associated to $\genjac_n^{(a,b)}$ and $P_n^{(b,b)}$ as
\begin{align}
x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x) 
\label{eqn:Hrecurrence} \\
y P_n^{(b,b)}(y) &= \delta_n^{(b)} P_{n+1}^{(b,b)}(y) + \gamma_n^{(b)} P_n^{(b,b)}(y) + \delta_{n-1}^{(b)} P_{n-1}^{(b,b)}(y),
\end{align}
where we note that \(\gamma_n^{(b)} = 0\), for \( n = 0,1,2,\dots\). We can use these to determine the 2D recurences for $\hdopnkab(x,y)$:

\begin{lemma}
$\hdopnk^{(a,b)}(x,y)$ satisfy the following 3-term recurrences:
\begin{align}
x \hdopnk^{(a,b)}(x,y) &= \alphaab_{n,k,1} \: \hdop_{n-1, k}^{(a,b)}(x, y) + \alphaab_{n,k,2} \: \hdop_{n, k}^{(a,b)}(x, y) + \alphaab_{n+1,k,1} \: \hdop_{n+1, k}^{(a,b)}(x, y), \\
y \hdopnk^{(a,b)}(x,y) &= \betaab_{n,k,1} \: \hdop_{n-1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,2} \: \hdop_{n-1, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,3} \: \hdop_{n, k-1}^{(a,b)}(x, y) + \betaab_{n,k,4} \: \hdop_{n, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,5} \: \hdop_{n+1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,6} \: \hdop_{n+1, k+1}^{(a,b)}(x, y),
\end{align}
for \((x,y) \in \Omega\), where
\begin{align}
\alphaab_{n,k,1} &:= \beta_{n-k-1}^{(a, b+k+\half)} \\
\alphaab_{n,k,2} &:= \alpha_{n-k}^{(a, b+k+\half)} \\
\betaab_{n,k,1} &:= {\delta_{k-1}^{(b)}}{|| \hdop_{n-1,k-1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjacnmk^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
\betaab_{n,k,2} &:= {\delta_{k}^{(b)}}{|| \hdop_{n-1,k+1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k-2}^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}} \\
\betaab_{n,k,3} &:= {\delta_{k-1}^{(b)}}{|| \hdop_{n,k-1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k+1}^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
\betaab_{n,k,4} &:= {\delta_{k}^{(b)}}{|| \hdop_{n,k+1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k-1}^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}} \\
\betaab_{n,k,5} &:= {\delta_{k-1}^{(b)}}{|| \hdop_{n+1,k-1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k+2}^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
\betaab_{n,k,6} &:= {\delta_{k}^{(b)}}{|| \hdop_{n+1,k+1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjacnmk^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}}. 
\end{align}
where $|| p(x,y) ||_{W^{(a,b)}}^2 := \int_\Omega p(x,y)^2 \: \Wab(x,y) \: d\mathbf{x}$ and $\ip<p(x), q(x)>_{\genjacw^{(a, b)}} := \int_\alpha^\beta p(x) \: q(x) \: x^a \: (1-x^2)^b \: dx$, i.e the 1D inner product using the $\{H^{(a,b)}_n\}$ family weight of $\genjacw^{(a, b)}$.

@@

\begin{proof}
Note that \(\ip<y \hdopnkab, \hdopmjab>_\Wab = 0\) for \(m < n-1\) (and for \(m > n+2\)). Thus for \(m = n-1, n, n+1,\) \(j = 0,\dots,m:\)
\begin{align}
\ip<y \hdopnkab, \hdopmjab>_\Wab &=  \iint_\Omega \hdopnkab(x,y) \: \hdopmjab(x,y) \: y \: \Wab(x,y) \: dy \: dx \\
&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(s) \: \genjacmmj^{(a, b+j+\half)}(s) \: s^a \: \rho(s)^{2b+k+j+2} \: ds \Big) \nonumber \\
& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: (1-t^2)^b \: dt \Big) \\
&= \begin{cases}
    	\delta_k^{(b)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}} \quad& \text{if } j = k+1 \\
	\delta_{k-1}^{(b)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k+\half)}} \quad& \text{if } j = k-1 \\
	0 & \text{otherwise}
      \end{cases}.
\end{align}

@@

Three-term recurrences lead to Jacobi operators that correspond to multiplication by $x$ and $y$. Define, for $n=0,1,2,\dots$: 
\begin{align}
\bighdopab_n := \begin{pmatrix}
		\hdopab_{n,0}(x,y) \\
		\vdots \\
		\hdopab_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1}, 
\quad \quad 
\bighdopab := \begin{pmatrix}
		\bighdopab_0 \\
		\bighdopab_1 \\
		\bighdopab_2 \\
		\vdots \\
	\end{pmatrix}
\end{align}
and set $J_x^{(a,b)}, J_y^{(a,b)}$ as the Jacobi matrices corresponding to
\begin{align}
J_x^{(a,b)} \: \bighdopab(x,y) = x \: \bighdopab(x,y), \quad J_y^{(a,b)} \: \bighdopab(x,y) = y \: \bighdopab(x,y).
\label{eqn:jacobimatricesdefinition}
\end{align}
The matrices $J_x^{(a,b)}, J_y^{(a,b)}$ act on the coefficients vector of a functions expansion in the $\{\hdopnkab\}$ basis. For example, let $a, b$ be general parameters and a function $f(x,y)$ defined on $\Omega$ be approximated by its expansion $f(x,y) = \bighdopab(x,y)^\top \mathbf{f}$. Then $x \: f(x,y)$ is approximated by $\bighdopab(x,y)^\top {J_x^{(a,b)\top}} \mathbf{f}$. In other words, ${J_x^{(a,b)\top}} \mathbf{f}$ is the coefficients vector for the expansion of the function $(x,y) \to x \: f(x,y)$ in the  $\{\hdopnkab\}$ basis.

@@

We can then obtain the matrix operator for the Laplacian \(\Delta\), that will take us from coefficients for expansion in the weighted space
$$
\bigWii(x,y) = \Wii(x,y) \: \bighdopii(x,y)
$$
to coefficients in the non-weighted space $\bighdopii(x,y)$. Note that this construction will ensure the imposition of the Dirichlet zero boundary conditions on $\Omega$. The matrix operator for the Laplacian we denote $\laplacewii$ acting on the coefficients vector is then given by
\begin{align}
    \laplacewii := D_x^{(0,0)} \: W_x^{(1,1)} + T^{(0,1)\to(1,1)} \: D_y^{(0,0)} \: T_W^{(1,0)\to(0,0)} \: W_y^{(1,1)}.
\end{align}
Importantly, this operator will have Banded-Block-Banded structure, and hence will be sparse, as seen in Figure \ref{fig:sparsity}.
